\section{Lecture 7: Nesterov’s accelerated gradient descent}

In this lecture, we present Nesterov's accelerated gradient descent for smooth convex functions. We will derive an algorithm that obtains a runtime of $\cO \left(\frac{\beta}{t^2}\right)$ for $\beta$-smooth functions. For functions which are $\alpha$-strongly convex, we will achieve a rate of $\cO \left(\exp\left( -\sqrt{\frac{\beta}{\alpha}} t \right) \right)$. The algorithm proceeds iteratively as follows. 
\begin{align*}
x_0 &= y_0 = z_0, \\
x_{k+1} &= \tau z_k + (1 - \tau) y_k \\
y_k &= x_k - \frac{1}{\beta} \nabla f(x_k) \\
z_k &= z_{k -1} - \eta \|\nabla f(x_k)\| \\
\end{align*}
The definitions are constructed so that we can apply bounds related to smooth and convex functions. We will simplify our proof of the runtime of the algorithm by ``restarting'' the procedure in the lemma several times; namely, after the error is reduced from $d$ to $\frac{d}{2}$, we run our algorithm again with a new $x_0^{'} = x_T$

\begin{lemma}
Suppose $\|x_0 - x^* \| \leq R$, $f$ attains its minimum at $x^*$ and is $\beta$-smooth. Also assume $f(x_0) -f(x^*) \leq d$.
Put $\eta = \frac{R}{\sqrt{d\beta}}$, and $\tau$ s.t. $\frac{1-\tau}{\tau} = \eta \beta$.\\
Then after $T = 4R\sqrt{\frac{\beta}{d}}$ steps, we have $$f(\bar{x})- f(x^*) \leq \frac{d}{2},$$ where $\bar{x} = \frac{1}{T} \sum_{k=0}^{T-1} x_k$.
\end{lemma}
% I think we formalize this argument later.
% What do we get if we ``restart'' the lemma several times?
% We get that after
% $$T = O\left(R \sqrt{\frac{\beta}{\epsilon}} + R \sqrt{\frac{\beta}{2\epsilon}} +...+ R \sqrt{\frac{\beta}{d}} \right) = O\left(\sqrt{\frac{\beta}{\epsilon}}\right)$$
% steps we get $\epsilon$-error.
% This is equivalent to stating that the error $\epsilon$ is $O\left(\frac{R\beta}{T^2}\right)$.

\begin{proof}
In lecture 2, we showed the following properties for smooth and convex functions.
\begin{equation}
f(y_k) - f(x_k) \leq -\frac{1}{2 \beta} \|\nabla f(x_k) \|^2. \label{Nesterov-pf_smoothness_and_convexity}
\end{equation}
By the ``Fundamental Theorem of Optimization" (see Lecture 2), we have
\begin{align}
\forall u: \ \eta  \langle \nabla f(x_{k+1}), z_k - u \rangle &= \frac{\eta^2 \|\nabla f(x_{k+1})\|^2 + \|z_k - u \|^2 - \|z_{k+1} - u \|^2}{2} \\
&\leq \frac{\eta^2}{2} \|\nabla f(x_{k+1})\|^2 + \|z_k - u \|^2 - \|z_{k+1} - u \|^2
\end{align}
Substituting the first equation yields
\begin{equation}
\eta \langle \nabla f(x_{k+1}, z_k - u \rangle \leq \eta^2 \beta (f(x_{k+1}) - f(y_{k+1})) + \|z_k - u\|^2 - \|z_{k+1} - u \|^2 \label{Nesterov-pf_eq3}
\end{equation}
\medskip
Besides,
\begin{align}
&\eta \langle \nabla f(x_{k+1}), x_{k+1} - u\rangle - \eta \langle \nabla f(x_{k+1}), z_k - u\rangle \nonumber\\
&= \eta \langle\nabla f(x_{k+1}), x_{k+1} - z_k\rangle \nonumber\\
&= \frac{1-\tau}{\tau}  \eta \langle\nabla f(x_{k+1}), y_k - x_{k+1}\rangle \nonumber\\
&\leq \frac{1-\tau}{\tau} \eta (f(y_k) - f(x_{k+1})) \text{ (by convexity)}. \label{Nesterov-pf_eq4}
\end{align}
\medskip
Combining  \eqref{Nesterov-pf_eq3} and \eqref{Nesterov-pf_eq4}, and setting $\frac{1-\tau}{\tau} = \eta \beta$ yield
\begin{equation*}
\forall u,\ \eta \langle\nabla f(x_{k+1}), x_{k+1} - u\rangle \leq \eta^2 \beta (f(y_k) - f(y_{k+1})) + \|z_k - u\|^2 - \|z_{k+1} - u\|^2.
\end{equation*}
This implies
\begin{equation*}
\eta T (f(\bar{x}) - f(x^*)) \leq \sum_{k=0}^T \eta \langle\nabla f(x_{k+1}), x_{k+1} - x^*\rangle\\
\leq \eta^2 \beta d + R^2,
\end{equation*}
which can be rewritten as
\begin{equation*}
f(\bar{x}) - f(x^*) \leq \frac{\eta \beta d}{T} + \frac{R^2}{\eta T}.
\end{equation*}
Setting $\eta = \frac{R}{\sqrt{\beta d}}$, this gives us
\begin{equation*}
f(\bar{x}) - f(x) \leq \frac{2 \sqrt{\beta d}}{T} R.
\end{equation*}
Therefore, for $T \geq 4 \sqrt{\frac{\beta}{d}} R$, we have
\begin{equation*}
f(\bar{x}) - f(x) \leq \frac{d}{2}.
\end{equation*}
\end{proof}

This lemma was initially proven by Allen-Zhu and Orecchia \cite{allen2014linear}. From the lemma, our analysis of the runtime of the algorithm naturally follows.

\begin{theorem}
\theoremlabel{smooth_nesterov}
 If a function $f$ is $\beta$-smooth and convex (not strongly convex), the lemma gives that after $T(d)=4R\sqrt{\frac{\beta}{d}}$ steps, 
\[f(\bar{x})- f(x^*)\leq \frac{d}{2}\qquad\text{, where } \bar{x} = \frac{1}{T}\sum_{\tau=0}^{T-1}x_k.\]
\end{theorem}
This means after $T(d)$ steps, Nesterov's accelerated gradient descent method reduces the function value error by half. For the complete Nesterov's method, after every $T(d)$ steps, let $\bar{x}$ be the new initial state. We restart the iterative updating algorithm from the beginning. %Then the error will keep decreasing until it is less then $\epsilon$.\\
\begin{proof}
[Proof of \theoremref{smooth_nesterov}.]
\indent By applying the Lemma $n$ times, we calculate the total iterations before $f(\bar{x}_{(n)})-f(x^*)\leq\frac{d}{2^{n}}\leq\epsilon$,
\begin{align*}
t(\epsilon) &= \cO \left(4R\sqrt{\frac{\beta}{d/2^{n-1}}} + 4R\sqrt{\frac{\beta}{d/2}} + 4R\sqrt{\frac{\beta}{d}}\right)\\
  &= \cO \left(4R\sqrt{\frac{\beta}{d/2^{n-1}}}\right)\\
  &= \cO \left(\sqrt{\frac{\beta}{\epsilon}}\right).
\end{align*}
\indent This can also be written as: 
\[\epsilon(t) = \cO \left(\frac{\beta}{t^2} \right)\]
\end{proof}

We can make a stronger statement for functions which are also strongly convex.

\begin{theorem}
\theoremlabel{smooth_strongly_convex}
If a function $f$ is $\beta$-smooth and $\alpha$-strongly convex, the Lemma gives that after constant $T=4\sqrt{\frac{2\beta}{\alpha}}$ steps,
\[\|\bar{x} - x^*\|^2 \leq \frac{1}{2}\|x_0 - x^*\|^2 \qquad\text{, where } \bar{x} = \frac{1}{T}\sum_{\tau=0}^{T-1}x_k.\]
\end{theorem}

\begin{proof}[Proof of \theoremref{smooth_strongly_convex}.]
\indent Using the same trick in 1), we apply the Lemma $n$ times before $\|\bar{x}_{(n)} - x^*\|^2 \leq \frac{1}{2^n}\|x_0 - x^*\|^2 \leq \epsilon$. Then the total iterations are:
\begin{align*}
t(\epsilon) &= \cO(nT)\\
            &= \cO \left (4\sqrt{\frac{2\beta}{\alpha}}\log_2 \left (\frac{\|x_0 - x^*\|^2}{\epsilon} \right) \right)\\
            &= \cO \left(-\sqrt{\frac{\beta}{\alpha}}\log_2\epsilon \right).
\end{align*}
\end{proof}

\indent In this case, Nesterov's accelerated gradient descent method yields linear convergence:
\[\epsilon(t) = 2^{-\Omega(t\sqrt{\frac{\alpha}{\beta}})}.\]

Table \ref{tab:nesterov} compares the bounds on error $\epsilon(t)$ while applying Nesterov's method and ordinary gradient descent method to different functions. Nesterov's accelerated gradient descent method has a faster convergence rate than the ordinary gradient descent method.

\begin{table}[h]
  \begin{center}
    \begin{tabular}{ | p{2cm} | c | c |}
      \hline
        & Nesterov's Method 
        & Ordinary GD Method \\ \hline
      $\beta$-smooth, convex 
      & \begin{tabular}{c}
        $\cO \left(\frac{\beta}{t^2}\right)$
        \rule[0cm]{0cm}{0.8cm}
        \end{tabular}
      & \begin{tabular}{c}
        $\cO \left(\frac{\beta}{t} \right)$
        \rule[0cm]{0cm}{0.8cm}
        \end{tabular} \\ \hline
      $\beta$-smooth, $\alpha$-strongly convex 
      & \begin{tabular}{c}
        $2^{-\Omega(t\sqrt{\frac{\alpha}{\beta}})}$
        \rule[0cm]{0cm}{1cm}
        \end{tabular} 
      & \begin{tabular}{c}
        $2^{-\Omega(t\frac{\alpha}{\beta})})$
        \rule[0cm]{0cm}{1cm}
        \end{tabular} \\ \hline
    \end{tabular}
  \end{center}
  \caption{Bounds on error $\epsilon$ as a function of number of iterations $t$ for different methods.}
  \label{tab:nesterov}
\end{table}

